Legion: Expressing Locality and Independence with Logical Regions

### 2012-05-02 doc/sc2012/circuit.tex
pass up to soop

Pointers to region values declare the region to which they point.  For
example, the definition of {\tt Wire} is parameterized on the region
{\tt rn} to which the {\tt Node} pointers in fields {\tt in\_nodes}
and {\tt out\_nodes} point, which guarantees that the two endpoints of
a wire are always in the same region.

>>>>

Pointers to region values declare the region to which they point.  For
example, the definition of {\tt Wire} is parameterized on the region
{\tt rn} to which the {\tt Node} pointers in fields {\tt in\_nodes}
and {\tt out\_nodes} point.
% which guarantees that the two endpoints of
%a wire are always in the same region.

### 2012-05-02 doc/sc2012/legion.tex
soop
pass up to soop

\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
\newcommand{\deref}{\ast}
\newcommand{\rread}[1]{\mbox{\em Read}(#1)}
\newcommand{\rwrite}[1]{\mbox{\em Write}(#1)}
\newcommand{\lca}[2]{#1 \sqcup #2}
\newcommand{\rleq}{\leq}
\newcommand{\interval}[1]{\mbox{\em interval}(#1)}
\newcommand{\context}[1]{\mbox{\em context}(#1)}
\newtheorem{theorem}{Theorem} 
\newtheorem{lemma}[theorem]{Lemma}

@@

Legion programs are written using logical regions, each of which
has one or more {\em physical instances} when the program executes.  Often
logical regions correspond to a single physical instance, but it is also common and useful
to have multiple physical copies of a logical region's contents in different parts of the machine 
to duplicate and distribute read-only data, to keep an old copy of a region to break a 
write-after-read dependency, or to allow independent, local reductions that are later combined.

@@

\item We use a {\em deferred execution} model that extracts parallelism
even from highly irregular, unbalanced computations.  Instead of blocking,
an operation $o$ that cannot immediately execute because it is dependent
on some other operation $o'$ waits on an event that occurs when $o'$ completes.
Events can be created for the completion of any operation in Legion, and 
the scheduler can build long chains of deferred operations.  In Section~\ref{sec:exp}
we show that dataflow-like deferred execution gives us a significant performance improvement
over bulk synchronous global synchronization primitives such as barriers.

@@

%\input{mapping_fig}

>>>>

\newcommand{\infrule}[2]{\displaystyle\frac{\displaystyle\strut{#1}}{\displaystyle\strut {#2}}}
\newcommand{\deref}{\ast}
\newcommand{\rread}[1]{\mbox{\em Read}(#1)}
\newcommand{\rwrite}[1]{\mbox{\em Write}(#1)}
\newcommand{\lca}[2]{#1 \sqcup #2}
\newcommand{\rleq}{\leq}
\newcommand{\interval}[1]{\mbox{\em interval}(#1)}
\newcommand{\context}[1]{\mbox{\em context}(#1)}
\newtheorem{theorem}{Theorem} 
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{observation}[theorem]{Observation}

@@

Legion programs are written using logical regions, which do not commit
to any particular placement of the data in the machine.  At runtime,
each logical reagion has one or more {\em physical instances}.  Often
logical regions correspond to a single physical instance, but it is also common and useful
to have multiple physical copies of a logical region's contents in different parts of the machine 
to duplicate and distribute read-only data, to keep an old copy of a region to break a 
write-after-read dependence, or to allow independent, local reductions that are later combined.

@@

%\item We use a {\em deferred execution} model that extracts parallelism
%even from highly irregular, unbalanced computations.  Instead of blocking,
%an operation $o$ that cannot immediately execute because it is dependent
%on some other operation $o'$ waits on an event that occurs when $o'$ completes.
%Events can be created for the completion of any operation in Legion, and 
%the scheduler can build long chains of deferred operations.  In Section~\ref{sec:exp}
%we show that dataflow-like deferred execution gives us a significant performance improvement
%over bulk synchronous global synchronization primitives such as barriers.

@@

### 2012-05-02 doc/sc2012/soop.tex
pass up to soop

If two children have a dependence between their region arguments, the
SOOP will schedule them to run serially to preserve the program's
sequential semantics.  If two child tasks have no dependence then
Legion can schedule them to execute out of order, in parallel.  Thus,
the first stage in the execution of a task is dependence analysis on
the logical regins used by tasks.

@@

Consider the first two subtasks spawned on line 32 of the circuit
simulation in Listing~\ref{lst:code_ex}, {\tt
calc\_new\_currents(pieces[0])} and {\tt
calc\_new\_currents(pieces[1])}.  The first of these tasks reads and
writes the private wires subregion {\tt pieces[0].rw\_pvt} in
exclusive mode, and reads in exclusive mode the node subregions {\tt
pieces[0].rn\_pvt}, {\tt pieces[0].rn\_shr}, and {\tt
pieces[0].rn\_ghost} (see line 39).  The second subtask must be
checked against the first for any dependences; the second subtask uses
{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
pieces[1].rn\_pvt}, {\tt pieces[1].rn\_shr}, and {\tt pieces[1].rn\_ghost} (read-only exclusive).
It may be helpful to refer to the region tree for nodes in Figure~\ref{sfig:part_fig:tree} (recall that the wires region tree consists of a single disjoint partition).  We reason about the dependences as follows:
\begin{enumerate}
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
a non-disjoint, or {\em aliased}, partition.
\item All the remaining pairs of regions used by the two tasks (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
do not overlap because they are in different region trees and have no common region ancestor.
\end{enumerate}
Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges.  Thus, these two subtasks are independent.

A closer look at this example shows that whether there is overlap
between two regions $r_1$ and $r_2$ can be determined by examining
their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  Note
that the region tree alternates between regions and partitions:
regions are always children of partitions and
partitions are always children of regions in the region tree.  There
are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
are in different region trees) or is a disjoint partition, then $r_1$
and $r_2$ are guaranteed to be disjoint.  If $\lca{r_1}{r_2}$ is an
aliased partition or a region (because $r_1$ and $r_2$ are in
different partitions of the region, as in cases 4 and 5
above), then $r_1$ and $r_2$ may not be disjoint.

Consider the subtasks {\tt distribute\_charge(pieces[0],dt)} and
{\tt update\_voltages[1]} in Listing~\ref{lst:code_ex}.  The former task
uses region {\tt pieces[0].rn\_ghost} and the latter uses region {\tt pieces[1].rn\_shr}.
The least common ancestor is the region of all shared nodes {\tt p\_nodes\_pvs[1]},
so these two regions are not disjoint.  Since both tasks write their respective subregions,
there is a dependence and the {\tt update\_voltages} task can only run after the
{\tt distribute\_charges} task has completed.  There is an important additional piece 
of information, as well.  Because the {\tt update\_voltages} task possibly reads the values
written by the {\tt distribute\_charges} task, the updates done by {\tt distribute\_charges} must
be reconciled on least common ancestor region {\tt p\_nodes\_pvs[1]}; our depedence analysis
tracks this information as well.

We can now describe the algorithm for the dependence analysis phase of
the SOOP.  The SOOP maintains a local region forest for each task $t$,
the roots of which are $t$'s region arguments and including all
partitions and subregions used within the task.  Dependence analysis
decorates the region tree with the subtasks of $t$, maintaining the
following invariant. Let $r'$ be a region argument to subtask $t'$,
and let $R$ be the set of regions $r''$ such that $r''$ is an argument
to a task $t''$ registered after $t'$ and $t''$ depends on $t'$ because
of overlap between $r''$ and $r'$.  Then $t'$ is listed on the region that
is the least common ancestor of the set $R \cup \{r'\}$.

We can identify all dependences between sibling tasks and place tasks
in the correct place in the region tree in amortized time ${\cal O}(nd)$,
where $n$ is the total number of region arguments to tasks and $d$ is the depth
of the region tree.  For region $r'$ an argument of task $t'$, the analysis
walks the unique path in the region forest from a root to $r'$.  For each node $m$
on the walk from the root to $r'$ the following actions are taken:
\begin{itemize}
\item Assume $m$ is a region with multiple child partitions, and let $m'$ be the child on the
path to $r'$.  Any task in a subtree of $m$ other than $m'$ may conflict with $t'$; these tasks
are all moved to $m$.

\item Assume $m$ is a non-disjoint partition, and let $m'$ be the child on the
path to $r'$.  Any task in a subtree of $m$ other than $m'$ may conflict with $t'$; these tasks
are all moved to $m$.

@@

\item Assume $m = r'$.  In this case any dependent tasks in $r'$'s subtree must be moved to $r'$.
\end{itemize}

@@

Thus, there are two parts to depedence analysis of a region argument
$r'$ to a task: the walk from a region root to $r'$ and the off-path work
to move dependent tasks to the least common ancestor on the path.
Clearly the walk to $r'$ is bounded by the depth of the region tree.
The off-path work can be made efficient by maintaining at each node a
record of which subtrees have tasks in them and with what priveleges,
enabling the off-path search to traverse directly and only to
dependent tasks.  The off-path work is proportional to how far a task
is moved back up the region tree.  Thus, for each region argument a
task is initially placed at some depth in the tree and subsequently
only moved up towards the root, so the overall work is (amortized)
${\cal O}(nd)$.  In practice, we have found this level of efficiency,
where dependence analysis is linear in the number of region arguments (and not
the square) and proportional to the depth of the region tree (and not its size)
to be very important in keeping the overhead of the SOOP low.

>>>>

If two children have a dependence between their region arguments, the
SOOP will schedule them to run serially to preserve the program's
sequential semantics.  If two child tasks have no dependence then
Legion can schedule them to execute out of order, in parallel.  Thus,
the first stage in the execution of a task is dependence analysis on
the logical regions used by tasks.

@@

Consider the first two subtasks spawned on line 32 of the circuit
simulation in Listing~\ref{lst:code_ex}, {\tt
calc\_new\_currents(pieces[0])} and {\tt
calc\_new\_currents(pieces[1])}.  The first of these tasks reads and
writes the private wires subregion {\tt pieces[0].rw\_pvt} in
exclusive mode, and reads in exclusive mode the node subregions {\tt
pieces[0].rn\_pvt}, {\tt pieces[0].rn\_shr}, and {\tt
pieces[0].rn\_ghost} (see line 39).  The second subtask must be
checked against the first for any dependences; the second subtask uses
{\tt pieces[1].rw\_pvt} (read/write exclusive) and {\tt
pieces[1].rn\_pvt}, {\tt pieces[1].rn\_shr}, and {\tt pieces[1].rn\_ghost} (read-only exclusive).
It may be helpful to refer to the region tree for nodes in Figure~\ref{sfig:part_fig:tree} (recall that the wires region tree consists of a single disjoint partition).  We reason about the dependences as follows:
\begin{enumerate}
\item {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rw\_pvt} do not overlap as they are subregions of a disjoint partition.
\item {\tt pieces[0].rn\_pvt} and {\tt pieces[1].rn\_pvt} do not overlap for the same reason.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_shr} do not overlap, also for the same reason.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_shr}, because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_shr} and {\tt pieces[1].rn\_ghost} may overlap because they are in two different
  partitions of the same region.
\item {\tt pieces[0].rn\_ghost} may overlap with {\tt pieces[1].rn\_ghost} because they are in different subregions of
a non-disjoint, or {\em aliased}, partition.
\item All the remaining pairs of regions used by the two tasks (e.g., {\tt pieces[0].rw\_pvt} and {\tt pieces[1].rn\_pvt})
do not overlap because they are in different region trees and have no common region ancestor.
\end{enumerate}
Cases 4-6 indicate a possible dependence, however in each case the regions are accessed with only read privileges.  Thus, the two subtasks are independent.

A closer look at this example shows that whether there is overlap
between two regions $r_1$ and $r_2$ can be determined by examining
their least common ancestor $\lca{r_1}{r_2}$ in the region tree.  Note
that the region tree alternates between regions and partitions:
regions are always children of partitions and
partitions are always children of regions.  There
are four cases.  If $\lca{r_1}{r_2}$ either does not exist (the regions
are in different region trees) or is a disjoint partition, then $r_1$
and $r_2$ are guaranteed to be disjoint.  If $\lca{r_1}{r_2}$ is an
aliased partition or a region (because $r_1$ and $r_2$ are in
different partitions of the region, as in cases 4 and 5
above), then $r_1$ and $r_2$ may not be disjoint.

Consider subtasks {\tt distribute\_charge(pieces[0],dt)} and
{\tt update\_voltages[1]} in Listing~\ref{lst:code_ex}.  The former task
uses region {\tt pieces[0].rn\_ghost} and the latter uses region {\tt pieces[1].rn\_shr}.
The least common ancestor is the region of all shared nodes {\tt p\_nodes\_pvs[1]},
so these two regions are not disjoint.  Since both tasks write their respective subregions,
there is a dependence and the {\tt update\_voltages} task can only run after the
{\tt distribute\_charges} task has completed.

We can now describe the algorithm for the dependence analysis phase of
the SOOP.  The SOOP maintains a local region forest for each task $t$,
the roots of which are $t$'s region arguments and including all
partitions and subregions used within the task.  Dependence analysis
decorates the region tree with the subtasks of $t$, maintaining the
following invariant. Let $r'$ be a region argument to subtask $t'$,
and let $R$ be the set of regions $r''$ such that $r''$ is an argument
to a task $t''$ registered after $t'$ and $t''$ depends on $t'$ because
of overlap between $r''$ and $r'$.  Then $t'$ is listed on the region that
is the least common ancestor of the set $R \cup \{r'\}$.  Moving $t'$ to an
ancestor of $r'$ has the effect of coarsening the dependences involving $t'$
for tasks registered in the future, and this can result in false dependences.
However, the loss of information can be shown to be small and, as we show next, the
benefit is that it dramatically speeds up dependence analysis overall.

We can identify all dependences between sibling tasks and place tasks
in the correct place in the region tree in amortized time ${\cal O}(nd)$,
where $n$ is the total number of region arguments to tasks and $d$ is the depth
of the region tree.  For region $r'$ an argument of task $t'$, the analysis
walks the unique path in the region forest from a root to $r'$.  For each node $m$
on the walk from the root to $r'$ the following actions are taken:
\begin{itemize}
\item Assume $m$ is a region with multiple child partitions, and let $m'$ be the child on the
path to $r'$.  Any task in a subtree of $m$ other than $m'$ on which $t'$ depends 
is moved to $m$.

\item Assume $m$ is a non-disjoint partition, and let $m'$ be the child on the
path to $r'$.  Any task in a subtree of $m$ other than $m'$ that $t'$ depends on is
moved to $m$.

@@

\item Assume $m = r'$.  In this case any dependent tasks in $r'$'s subtree are moved to $r'$.
\end{itemize}

@@

Thus, there are two parts to depedence analysis of a region argument
$r'$ to a task: the walk from a region root to $r'$ and the off-path work
to move dependent tasks to the least common ancestor on the path.
Clearly the walk to $r'$ is bounded by the depth of the region tree.
The off-path work can be made efficient by maintaining at each node a
record of which subtrees have tasks in them and with what priveleges,
enabling the off-path search to traverse directly and only to
dependent tasks.  The off-path work is proportional to how far a task
is moved back up the region tree.  Thus, for each region argument a
task is initially placed at some depth in the tree and subsequently
only moved up towards the root, so the overall work is (amortized)
${\cal O}(nd)$.  In practice, we have found this level of efficiency,
where dependence analysis is linear in the number of region arguments (and not
the square) and proportional to the depth of the region tree (and not its size)
to be very important in minimizing the overhead of the SOOP.
